Synthesising Complex Termination Arguments

### 2014-06-05 papers/termination/paper.tex
Update paper with current strategy

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
(lexicographic) linear and non-linear ones, 
for programs with both bit-vectors and floats. Our technique can handle programs with non-linear operations, e.g. logical and.
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem). 
As a direct consequence, our technique is semantic in nature and does not
dependent on the size of the analysed program or its number of variables. 
However, its efficiency does depend on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?). 
%We empirically observe that most programs have
%ranking functions with low Kolmogorov complexity (see Section ?).

@@

The main contributions of our paper are highlighted below:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.  
\end{itemize}

@@

\section{Related Work}
Error 404.

@@

@@

\section{Formalism}
In this section, we fix a model of computation, describe its semantics and
define the syntax of a language we will work over.

@@

\section{Combinatorics of Termination}

\begin{itemize}
 \item There are a TON of programs (way more than you'd expect).
 \item There are a TON of terminating programs, and for our model of computation we can count
  how many (the Chaitin constant).
 \item There are a TON of ranking functions (way more than you'd expect, but not many as a
  fraction of programs).
 \item There are not many linear functions.
 \item Most terminating programs don't admit linear ranking functions.
 \item The Curry-Howard isomorphism
 \item Kolmogorov complexity is relevant for understanding termination proofs.
\end{itemize}

@@

\begin{definition}[Second-order Termination Formula]
\begin{align*}
 \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

@@

\section{Solving the Synthesis Formula}
\input{kalashnikov}

@@

We implemented our fully automatic program synthesis procedure in the \textsc{Kalashnikov} tool.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{pifont}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage{pgf}
\usepackage[noend]{algpseudocode}
\usepackage{algorithm}
\usepackage{multicol}
\usepackage{appendix}
\usepackage{caption}
\DeclareCaptionType{copyrightbox}
\usepackage{subfig}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, automata, shapes}

@@

Because most state-of-the-art termination procedures use disjunctive well foundedness arguments, they require the use of
complex safety provers that are able to reason about unbounded unfoldings of a transition relation.  In contrast to this,
we are able to build monolithic termination arguments that only need ``local'' reasoning -- our ranking functions
can be proved sufficient by considering just a single unfolding of a loop.

Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
(lexicographic) linear and non-linear ones, 
for programs with both bit-vectors and floats. Our technique can handle programs with non-linear operations, e.g. logical and.
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem). 
As a direct consequence, our technique is semantic in nature and does not
dependent on the size of the analysed program or its number of variables. 
However, its efficiency does depend on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).

By phrasing the termination problem as a single rank function synthesis problem, we observe that
we can make the termination problem as hard as finding ranking functions, rather than as hard as
solving a difficult ``double reachability'' problem (cf. Terminator).

\begin{figure}
 \begin{tabular}{|ll||c|c|}
 \hline
  & & \multicolumn{2}{c|}{Program} \\
  & & Linear & Non-Linear \\
  \hline
  \hline
  \multirow{2}{*}{Ranking Functions} & Linear & \checkmark \xmark & \checkmark \\
   & Non-Linear & \checkmark & \checkmark \\
   \hline
 \end{tabular}

 \caption{Legend: \xmark = state-of-the art can handle, \checkmark = we can handle\label{fig:handletable}}
\end{figure}

%We empirically observe that most programs have
%ranking functions with low Kolmogorov complexity (see Section ?).

@@

The main contributions of our paper are highlighted below:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
we don't do a DWF proof).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

@@

\section{Related Work}
Primarily divided into structural (syntactic) vs. semantic.  For example, terminationg of term rewriting systems
is largely structural, whereas DWF analysis (cf. Terminator) is more semantic -- it just deals
with transition relations, but uses some amount of syntax in that it identifies loops \& looping
paths.

@@

\subsection{Disjunctive Well-Foundedness}
Terminator uses disjunctive well foundedness arguments, which can be found
piece by piece, but require reasoning about the transitive closure of a transition
relation.  Because we build a monolithic well foundedness argument, we
do not need to reason about the full transitive closure, and can instead do our
reasoning over a single step of the transition relation.  This means that we
can get away with very simple model checking methods, rather than requiring
a full-blown safety prover.

@@

\section{Combinatorics of Finite Termination}
In this section, we fix a model of computation, describe its semantics and
define the syntax of a language we will work over.

@@

\subsection{Counting Programs}
\begin{itemize}
 \item There are a TON of programs (way more than you'd expect).
 \item There are a TON of terminating programs, and for our model of computation we can count
  how many (the Chaitin constant).
 \item There are a TON of ranking functions (way more than you'd expect, but not many as a
  fraction of programs).
 \item There are not many linear functions.
 \item Most terminating programs don't admit linear ranking functions.
 \item The Curry-Howard isomorphism
 \item Kolmogorov complexity is relevant for understanding termination proofs.
\end{itemize}

@@

\begin{figure*}
\begin{definition}[Second-order Termination Formula]
\begin{align*}
 \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
 \begin{align*}
  \exists R, I . \forall x, x' . & P(x) \rightarrow I(x) ~ \wedge \\
                                 & c(x) \wedge I(x) \wedge T(x, x') \rightarrow I(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
 \begin{align*}
  \exists R, I_1, I_2 . \forall x, x' . & P(x) \rightarrow I_1(x) ~ \wedge \\
                                        & I_1(x) \wedge L_1(x, x') \rightarrow I_1(x') ~ \wedge \\
                                        & I_1(x) \wedge \lnot c_1(x) \rightarrow I_2(x) ~ \wedge \\
                                        & c(x) \wedge I_2(x) \wedge L_2(x, x') \rightarrow I_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}
\end{figure*}

@@

\section{Solving the Synthesis Formula}
We can solve the synthesis problem using any off-the-shelf program synthesiser.  If you don't
have a program synthesiser kicking around, we present the design of one synthesiser
that can be used to solve the problem.

\input{kalashnikov}

@@

We proved termination for a bunch of programs, see Fig.~\ref{fig:linear} and Fig.~\ref{fig:nonlinear}.
